---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Shana Shapiro Section #1"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Monday, February 28 at 7:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
getwd()
library('tidyverse')
library('corrplot')
library('agricolae')
library('ggplot2')
library("cowplot")
library('lubridate')
library('dplyr')
#lakes <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv") #while editing
lakes <- read.csv("../Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv") #while knitting/rendering 
lakes$sampledate <- as.Date(lakes$sampledate, format = "%m/%d/%y")

#2
mytheme <- theme_light(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        plot.title = element_text(face = "bold",size=12),
        axis.ticks = element_line(colour="grey70", size = 0.2),
        panel.grid.major = element_line(colour="grey70", size = 0.2),
        panel.grid.minor = element_blank())
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: There is no difference in mean lake temperature recorded during July as depth changes across all lakes 
Ha: Mean lake temperature recorded during July changes with depth across all lakes 


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 째C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
lakes.wrangle <- lakes %>%
  filter(month(sampledate) %in% 07) %>%
  select(`lakename`, `year4`, `daynum`, `depth`, `temperature_C`) %>%
  na.omit(lakes.wrangle)
  
#5
lakes.wrangle.vis <- ggplot(lakes.wrangle, aes(x=depth, y=temperature_C)) +
  geom_point() +
  geom_smooth(method = lm, color = "green") + 
  xlim(0,15) + 
  ylim(0,35) +
  ylab(expression("Temperature (째C)")) + 
  xlab(expression("Depth (m)")) +
  mytheme
print(lakes.wrangle.vis)
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: As depth increases, temperature decreases. Temperatures appear to decrease quickly and then level off suggesting that the trend is not entirely linear, though there appears to be a strong decreasing trend in the relationship.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
lakes.reg <- lm(data = lakes.wrangle, temperature_C ~ depth)
summary(lakes.reg)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: As depth increases, there is a significant difference in temperature. 73.87% of the variability in temperatures is explained by changes in depth. There are 1 degree of freedom with a p-value of <2.2e-16. For every 1m change in depth, the temperature is predicted to decrease by 1.946째C


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
lake.AIC <- lm(data = lakes.wrangle, temperature_C ~ year4 + depth + daynum)
summary(lake.AIC)

#10
step(lake.AIC)
lake.model <- lm(data = lakes.wrangle, temperature_C ~ year4 + depth + daynum)
summary(lake.model)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of explanatory variables includes year4, depth, and daynum in the multiple regression. In conducting the step function, only one model was suggested with the three variables. Overall, the model explains 74.12% of the observed variance (p < 2.23e-16, DF = 3). This is a slight improvement on the initial model using only depth as the explanatory variable, where 73.87% of the variance was explained. 

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
lakes.aov <- aov(data=lakes.wrangle, temperature_C ~ lakename)
summary(lakes.aov)
lakes.lm <- lm(data=lakes.wrangle, temperature_C ~ lakename)
summary(lakes.lm)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: There is a significant difference in mean temperature among lakes. According to the one-way analysis of variance, we can reject the null hypothesis that there is no difference in temperature in the month of July according to lake (p<2e-16, df = 8).



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
lakes.bylake <- ggplot(lakes.wrangle, aes(x=depth, y=temperature_C, color = lakename)) + 
  geom_point(alpha = 0.5) +
  geom_smooth(aes(group=lakename), method="lm", color = "black", size=0.5, se = FALSE) +
  xlim(0,15) + 
  ylim(0,35) +
  ylab(expression("Temperature (째C)")) + 
  xlab(expression("Depth (m)")) +
  mytheme +
  scale_color_manual(values = c("Paul Lake" = "coral3",
                                "Peter Lake" = "dodgerblue4",
                                "Tuesday Lake" = "darkorange4",
                                "East Long Lake" = "darkseagreen4",
                                "West Long Lake" = "mediumpurple",
                                "Central Long Lake" = "mediumturquoise", 
                                "Hummingbird Lake" = "midnightblue", 
                                "Crampton Lake" = "orange4", 
                                "Ward Lake" = "violetred4"),"Lake Name")
print(lakes.bylake)
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(lakes.aov)
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: If the p-value is > 0.05, we cannot reject the null hypothesis that there is no difference in mean temperature of the lake. By that logic, the lakes statistically have the same mean temperature. Paul Lake and Ward Lake  have the same mean temperature. Every lake has an interaction with another lake where the p-value > 0.05. Therefore there each lake is statistically the same as at least one other lak and there is no single lake that is statistically distinct from all of the other lakes. 

NO (Crampton, Central, Ward, Hummingbird, East, Tuesday, West, Peter, Paul, )


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: If only investigating Peter Lake and Paul lake, another test of measuring distinct mean temperatures includes a two-sample t test, which checks if the means of two samples are equivalent.  


